---
title: 'Lab Session: Vector Autoregression (VAR) and Structural VAR (SVAR)'
author: "Econometrics Lab"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
  pdf_document:
    toc: true
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6,
  fig.align = "center"
)
```

# Introduction

## What Is a VAR?

A **Vector Autoregression (VAR)** is a system of equations where each variable is regressed on its own lags and the lags of all other variables in the system. It generalizes the univariate AR model to a multivariate setting.

For a bivariate VAR(1) with variables $y_t$ and $x_t$:

$$
\begin{aligned}
y_t &= c_1 + a_{11} y_{t-1} + a_{12} x_{t-1} + u_{1t} \\
x_t &= c_2 + a_{21} y_{t-1} + a_{22} x_{t-1} + u_{2t}
\end{aligned}
$$

In matrix form: $\mathbf{Y}_t = \mathbf{c} + \mathbf{A}_1 \mathbf{Y}_{t-1} + \mathbf{u}_t$

**Key features:**

-   Every variable is endogenous
-   No contemporaneous terms on the RHS (only lags)
-   The error terms $u_t$ may be correlated across equations

## Why VAR?

-   Captures dynamic interdependencies among multiple time series
-   Useful for forecasting, Granger causality testing, and impulse response analysis
-   Minimal structural assumptions required

## Reduced-Form vs. Structural

|   | Reduced-Form VAR | Structural VAR (SVAR) |
|------------------------|------------------------|------------------------|
| **Contemporaneous effects** | Hidden in error covariance | Explicitly modeled |
| **Errors** | Correlated ($\Sigma_u$) | Orthogonal structural shocks ($\varepsilon_t$) |
| **Identification** | Not needed | Requires restrictions |
| **Impulse responses** | Not structurally interpretable | Economically meaningful |

------------------------------------------------------------------------

# Setup: Packages and Data

```{r packages}
# Install packages if needed (uncomment the lines below on first run)
# install.packages(c("vars", "svars", "urca", "tidyverse", "tseries", "lmtest"))

library(vars)       # VAR estimation and tools
library(svars)      # Structural VAR identification
library(urca)       # Unit root / cointegration tests
library(tseries)    # Additional time series tests
library(lmtest)     # Granger causality
library(tidyverse)  # Data manipulation and plotting
```

## Load the Data

We use the **Canada** dataset built into the `vars` package. It contains four quarterly macroeconomic time series for Canada (1980Q1–2000Q4):

-   `e` — Employment
-   `prod` — Labor productivity
-   `rw` — Real wage
-   `U` — Unemployment rate

```{r load-data}
data("Canada")

# Inspect the data
cat("Class:", class(Canada), "\n")
cat("Dimensions:", dim(Canada), "\n")
cat("Time range:", start(Canada), "to", end(Canada), "\n\n")

head(Canada)
```

## Visualize the Series

Always plot your data before modeling!

```{r plot-data}
# Convert to data frame for ggplot
canada_df <- data.frame(
  date = as.Date(time(Canada)),
  as.data.frame(Canada)
)

canada_long <- canada_df %>%
  pivot_longer(cols = -date, names_to = "variable", values_to = "value")

ggplot(canada_long, aes(x = date, y = value)) +
  geom_line(color = "steelblue", linewidth = 0.8) +
  facet_wrap(~ variable, scales = "free_y", ncol = 2) +
  labs(title = "Canadian Macroeconomic Variables (1980Q1 – 2000Q4)",
       x = "Date", y = "Value") +
  theme_minimal(base_size = 13)
```

> **Discussion:** Do the series look stationary? Which ones appear to have trends or unit roots?

------------------------------------------------------------------------

# Preliminary Analysis

## Unit Root Tests

Before estimating a VAR, we need to determine the order of integration for each series.

```{r unit-root}
# ADF test for each variable
cat("=== Augmented Dickey-Fuller Tests ===\n")
cat("H0: The series has a unit root (non-stationary)\n\n")

for (var_name in colnames(Canada)) {
  adf_result <- adf.test(Canada[, var_name])
  cat(sprintf("%-6s: ADF stat = %6.3f, p-value = %.4f  %s\n",
              var_name, adf_result$statistic, adf_result$p.value,
              ifelse(adf_result$p.value < 0.05, "[STATIONARY]", "[NON-STATIONARY]")))
}
```

> **Interpretation:** If a series is non-stationary (fail to reject H0), we might consider:
>
> 1.  Differencing the data and estimating a VAR in differences
> 2.  Estimating a VAR in levels (valid if the VAR is stable — see Lütkepohl 2005)
> 3.  Using a VECM if cointegration is present
>
> For this lab, we proceed with the data **in levels**, which is a common practice in applied macro when the focus is impulse responses.

------------------------------------------------------------------------

# VAR Estimation

## Step 1: Lag Order Selection

Choosing the right number of lags is critical. Too few → omitted dynamics and autocorrelated residuals. Too many → loss of degrees of freedom and overfitting.

```{r lag-selection}
# VARselect tests up to lag p_max
lag_select <- VARselect(Canada, lag.max = 8, type = "const")

cat("=== Information Criteria by Lag Order ===\n\n")
print(lag_select$selection)

cat("\n--- Full IC Table ---\n")
print(round(lag_select$criteria, 2))
```

The most commonly used criteria:

-   **AIC** (Akaike): Tends to choose more lags — good for capturing dynamics
-   **SC/BIC** (Schwarz/Bayesian): Penalizes more heavily — parsimonious models
-   **HQ** (Hannan-Quinn): A middle ground

## Step 2: Estimate the VAR

```{r estimate-var}
# Estimate VAR with the lag order chosen by BIC
p_bic <- lag_select$selection["SC(n)"]
cat("Selected lag order (BIC):", p_bic, "\n\n")

var_model <- VAR(Canada, p = p_bic, type = "const")

# Summary of one equation (e.g., the Employment equation)
cat("=== Employment Equation ===\n")
summary(var_model$varresult$e)
```

## Understanding the Output

```{r var-coefficients}
# Extract all coefficient matrices
cat("=== VAR Coefficient Matrices ===\n\n")

# For each lag, show the A matrix
for (i in 1:p_bic) {
  cat(sprintf("--- A_%d (Lag %d coefficients) ---\n", i, i))
  # Extract coefficients for lag i
  coef_matrix <- sapply(var_model$varresult, function(eq) {
    coefs <- coef(eq)
    # Get coefficients for this lag (variable names end in .li)
    lag_pattern <- paste0("\\.l", i, "$")
    coefs[grep(lag_pattern, names(coefs))]
  })
  print(round(coef_matrix, 4))
  cat("\n")
}

# Residual covariance matrix (this is the key object linking VAR to SVAR)
cat("--- Residual Covariance Matrix (Sigma_u) ---\n")
print(round(summary(var_model)$covres, 6))

cat("\n--- Residual Correlation Matrix ---\n")
print(round(summary(var_model)$corres, 4))
```

> **Key insight:** The off-diagonal elements in the correlation matrix tell us how much the reduced-form shocks co-move. If they are large, ordering matters a lot for impulse responses!

------------------------------------------------------------------------

# Diagnostic Checks

A well-specified VAR should have residuals that are approximately white noise.

## Serial Correlation Test

```{r serial-corr}
# Portmanteau test (multivariate Ljung-Box)
serial_test <- serial.test(var_model, lags.pt = 16, type = "PT.asymptotic")

cat("=== Portmanteau Test for Serial Correlation ===\n")
cat("H0: No serial correlation up to lag 16\n")
cat(sprintf("Chi-sq = %.3f, df = %d, p-value = %.4f\n",
            serial_test$serial$statistic,
            serial_test$serial$parameter,
            serial_test$serial$p.value))
cat(ifelse(serial_test$serial$p.value > 0.05,
           "=> Fail to reject H0: No evidence of serial correlation.\n",
           "=> Reject H0: Serial correlation detected! Consider adding more lags.\n"))
```

## Stability Check

The VAR is stable (stationary) if all eigenvalues of the companion matrix lie inside the unit circle.

```{r stability}
# Check stability
roots_var <- roots(var_model)

cat("=== Stability Check ===\n")
cat("Eigenvalue moduli of companion matrix:\n")
print(round(roots_var, 4))
cat(sprintf("\nMax modulus: %.4f\n", max(roots_var)))
cat(ifelse(all(roots_var < 1),
           "=> All roots inside unit circle: VAR is STABLE.\n",
           "=> WARNING: Roots outside unit circle — VAR is NOT stable!\n"))
```

# Granger Causality

Granger causality tests whether lags of one variable help predict another, beyond the latter's own lags.

**Formally:** $x$ Granger-causes $y$ if past values of $x$ contain information useful for predicting $y$, beyond what is contained in past values of $y$ alone.

```{r granger}
cat("=== Granger Causality Tests ===\n")
cat("H0: [column variable] does NOT Granger-cause [row variable]\n\n")

vars_list <- colnames(Canada)

# Create a summary table
granger_results <- data.frame(
  Cause = character(),
  Effect = character(),
  F_stat = numeric(),
  p_value = numeric(),
  Conclusion = character(),
  stringsAsFactors = FALSE
)

for (cause_var in vars_list) {
  gc <- causality(var_model, cause = cause_var)
  # The test is: cause_var Granger-causes the remaining variables
  f_val <- gc$Granger$statistic
  p_val <- gc$Granger$p.value
  cat(sprintf("%s => {others}: F = %.3f, p = %.4f  %s\n",
              cause_var, f_val, p_val,
              ifelse(p_val < 0.05, "[SIGNIFICANT]", "[not significant]")))
}
```

> **Caution:** Granger causality ≠ true causality. It only measures predictive content. A variable can Granger-cause another due to a common omitted driver.

------------------------------------------------------------------------

# Impulse Response Functions (IRFs)

## Reduced-Form (Cholesky) IRFs

The most common approach uses a **Cholesky decomposition** of the residual covariance matrix. This imposes a recursive (triangular) ordering: the first variable is not contemporaneously affected by shocks to any other variable; the second is affected only by the first, and so on.

**The ordering matters!** Variables ordered first are treated as more "exogenous."

```{r irf-cholesky}
# Default ordering: e, prod, rw, U (as they appear in the data)
cat("Variable ordering:", paste(colnames(Canada), collapse = " -> "), "\n")
cat("Interpretation: e is most exogenous, U is most endogenous\n\n")

# IRF: shock to productivity -> all variables
irf_prod <- irf(var_model,
                impulse = "prod",
                response = c("e", "prod", "rw", "U"),
                n.ahead = 20,
                boot = TRUE,
                ci = 0.95,
                runs = 500)

plot(irf_prod, main = "Impulse Responses to a Productivity Shock (Cholesky)")
```

```{r irf-employment}
# IRF: shock to employment -> all variables
irf_e <- irf(var_model,
             impulse = "e",
             response = c("e", "prod", "rw", "U"),
             n.ahead = 20,
             boot = TRUE,
             ci = 0.95,
             runs = 500)

plot(irf_e, main = "Impulse Responses to an Employment Shock (Cholesky)")
```

## How Sensitive Is the Ordering?

Let's compare two different orderings to see if the results change materially.

```{r ordering-comparison}
# Ordering 1: e, prod, rw, U (original)
var_order1 <- VAR(Canada[, c("e", "prod", "rw", "U")], p = p_bic, type = "const")

# Ordering 2: U, rw, prod, e (reversed)
var_order2 <- VAR(Canada[, c("U", "rw", "prod", "e")], p = p_bic, type = "const")

irf1 <- irf(var_order1, impulse = "prod", response = "e", n.ahead = 20,
            boot = TRUE, runs = 500)
irf2 <- irf(var_order2, impulse = "prod", response = "e", n.ahead = 20,
            boot = TRUE, runs = 500)

# Compare
par(mfrow = c(1, 2))
plot(irf1, main = "Order: e → prod → rw → U")
plot(irf2, main = "Order: U → rw → prod → e")
par(mfrow = c(1, 1))
```

> **Takeaway:** If results change substantially with ordering, you should use a **Structural VAR** with theory-motivated identification instead of relying on arbitrary Cholesky orderings.

------------------------------------------------------------------------

# Forecasting with VAR

```{r forecast}
# Generate forecasts
var_forecast <- predict(var_model, n.ahead = 8, ci = 0.95)

# Numerical forecast for employment
cat("=== 8-Quarter Ahead Forecast for Employment ===\n")
print(round(var_forecast$fcst$e, 3))
```

------------------------------------------------------------------------
